Is this normal? A new projection pursuit index to assess a sample against a multivariate null distribution

### 2024-07-24 paper.tex
Formatting changed to tandf template

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand\bD{$1\text{-}D$}
\newtheorem*{theorem}{Theorem}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

@@

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{\bf Is this normal? A new projection pursuit index to assess a
sample against a multivariate null distribution}
\author{
Annalisa Calvi\\
School of Mathematics, Monash University\\
and\\Ursula Laa\\
Institute of Statistics, University of Natural Resources and Life
Sciences, Vienna\\
and\\Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
}
\maketitle

\bigskip
\bigskip
\begin{abstract}
Many data problems contain some reference or normal conditions, upon
which to compare newly collected data. This scenario occurs in data
collected as part of clinical trials to detect adverse events, and or
for measuring climate change against historical norms. The data is
typically multivariate, and often the normal ranges are specified by a
multivariate normal distribution. The work presented in this paper
develops methods to compare the new sample against the reference
distribution with high-dimensional visualisation. It uses a projection
pursuit guided tour to produce a sequence of low-dimensional projections
steered towards those where the new sample is most different from the
reference. A new projection pursuit index is defined, and the drawing of
the projected ellipse is computed analytically. The methods are
implemented in the R package, \texttt{tourr}.
\end{abstract}

\noindent%
{\it Keywords:} exploratory data analysis, statistical
graphics, dimension reduction, high-dimensional visualisation, climate
change, clinical trials
\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!

\section{Introduction}\label{introduction}

Linear projections are useful in many aspects of statistical analysis of
multivariate data, and especially useful for visualising the data. A
linear projection provides a dimension reduction while maintaining
interpretability. For example, a biplot \citep{Ga71, GH96} shows the
structure creating the maximum variance in the data, and also
visualizing the projection matrix to learn which variables contribute to
it. We might find clusters of outliers that were hiding in high
dimensions.

More generally, projection pursuit \citep{FT74, JS87, Hu85} defines a
quantitative criterion for the interestingness of a projection (a
projection pursuit index), and searches the space of possible
projections for the most interesting one to display. We can also define
sequences of interpolated linear projections to better understand a
multivariate distribution. Animating a randomly selected interpolated
sequence of linear projections is called a grand tour
\citep{As85, BA86b, BCAH05, CLBW06, tours2022}. The combination of these
two approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour \citep{cook1995}.

@@

To define a measure of an interesting projection is to maximize the
average Mahalanobis distance \citep{mahalanobis} in the projection for a
subset of points, \(W\). The set of points could be chosen in different
ways, but the default is those that are outside the specified ellipsoid
in \pD{}. Alternatives could be to select a set of observations with the
largest Mahalanobis distance, manually select observations or possibly a
group of points identified via clustering of the extremes.

@@

This is implemented in the \texttt{tourr} \citep{tourr, tourr-cran}
package, where the projected ellipsoid can be drawn for each projection.
The guided tour will take arguments specifying the data, and the null
variance-covariance matrix.

@@

This example is motivated by a problem posed during consulting with a
pharmaceutical company, but the data shown here is simulated, simply to
illustrate the application. Liver function tests commonly provide
measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase
(GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and
alanine aminotransferase (ALT). There are normal ranges on these
measurements reported by \citet{lib-med-liver-norms} and listed in
Table~\ref{tbl-liver-norms}.

@@

Figure~\ref{fig-liver} illustrates two examples. The first is similar to
the consulting project. A sample of liver test scores for new patients
was provided in order to examine their values relative to the normal
range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows
a projection of this sample relative to the normal range. Three of the
patients are outside the confidence ellipse but all of the patients are
located away from the mean. What has been typical in the past is to
compute normal values based on tests of healthy young males. The samples
provided for the project were all recorded on women. We see that this
sample has slightly lower ALT and ALP, which is consistent with what is
reported in \citet{lib-med-liver-norms}.

@@

This example is motivated by the weather data example from
\citet{filzmoser2018}. We illustrate using the anomaly index to compare
potential outliers with a reference normal distribution that is derived
using robust methods on the original data. The data contains 16
numerical measurements that are averages across the three summer months
June, July and August, reported for 68 years (1955 - 2022) (see
\citet{filzmoser2018} for more details).

To estimate the underlying normal distribution the data is first
centered and scaled using the median and the median absolute deviation
(MAD), before applying the minimum covariance determinant (MCD)
estimator \citep{rousseeuw1985multivariate} using the implementation in
\citet{robustbase}. The MCD estimates for the mean vector and the
variance-covariance matrix are then used to define the reference normal
distribution.

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in \citet{filzmoser2018},
we will further separate the outlying points into clusters based on
similarity in direction. The similarity is computed by first normalizing
observations to have length \(1\) and then apply k-means clustering with
Euclidean distance. We use the Dunn index \citet{dunn}, computed using
\citet{fpc}, to select the preferred number of clusters, \(k=5\).

@@

To compare our results to what is obtained by outlier detection methods,
we reproduce one of the results from \citet{filzmoser2018} in
Figure~\ref{fig-shapley}. On top we can see the cluster assignments for
all points that we have identified as outliers using the
\(p\)-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from \citet{filzmoser2018} also indicates
unusually low values for the minimum teperature, while there is
disagreement between outlyingness in other variables. Note that compared
to the years 1965 and 1966 they have unusually low minimum temperature
while having average values for the number of days with maximum
temperature above 25°C, and this contrast is also captured in the
projection.

@@

\caption{\label{fig-shapley}Reproduction of the results from
\citet{filzmoser2018}, showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment at the top.}

@@

The work is related to outlier detection methods \citep[see,][for
example]{filzmoser2018}, particularly those that use robust statistics,
as illustrated in the weather example. Methods discussed in
\citet{directional-outlyingness} and \citet{donoho1982} describe
detecting outlyingness using projections of the data. When these
projections are collected and used as a large set, they define a
high-dimensional confidence region of an indeterminate shape. There is
no analytical definition. However, we could explore the data relative to
the high-dimensional shapes using the same approach described in
Section~\ref{sec-background}. You would generate points on the surface
of the irregular shape, overlay the data on this to visualise with a
tour.

@@

Another potential direction of this work is with model-based clustering
using Gaussian mixtures \citep{FR02}, where ellipses form the basis of
the model. Currently, the \texttt{mclust} software \citep{mclust} has
the capacity to show \gD{} ellipses for two variables, or axis parallel
\gD{} ellipses for \pD{}. The equations developed here for drawing the
\gD{} projection of a \pD{} ellipse would provide more versatile
visualisation for examining the model-based clustering fits.

@@

  \bibliography{bibliography.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{interact}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{bm}
\def\tightlist{}
\usepackage{setspace}
\newcommand\pD{$p\text{-}D$}
\newcommand\kD{$k\text{-}D$}
\newcommand\dD{$d\text{-}D$}
\newcommand\gD{$2\text{-}D$}
\newcommand\bD{$1\text{-}D$}
\newtheorem*{theorem}{Theorem}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\title{Is this normal? A new projection pursuit index to assess a sample
against a multivariate null distribution}
\author{Annalisa Calvi$\textsuperscript{1}$, Ursula
Laa$\textsuperscript{2}$, Dianne Cook$\textsuperscript{3}$}

\thanks{CONTACT: Annalisa
Calvi. Email: \href{mailto:annalisa.calvi@monash.edu}{\nolinkurl{annalisa.calvi@monash.edu}}. Ursula
Laa. Email: \href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}. Dianne
Cook. Email: \href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} School of Mathematics, Monash
University,  \\ \textsuperscript{2} Institute of Statistics, University
of Natural Resources and Life Sciences,
Vienna,  \\ \textsuperscript{3} Department of Econometrics and Business
Statistics, Monash University,  
\begin{abstract}
Many data problems contain some reference or normal conditions, upon
which to compare newly collected data. This scenario occurs in data
collected as part of clinical trials to detect adverse events, and or
for measuring climate change against historical norms. The data is
typically multivariate, and often the normal ranges are specified by a
multivariate normal distribution. The work presented in this paper
develops methods to compare the new sample against the reference
distribution with high-dimensional visualisation. It uses a projection
pursuit guided tour to produce a sequence of low-dimensional projections
steered towards those where the new sample is most different from the
reference. A new projection pursuit index is defined, and the drawing of
the projected ellipse is computed analytically. The methods are
implemented in the R package, \texttt{tourr}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
exploratory data analysis\sep statistical graphics\sep dimension
reduction\sep high-dimensional visualisation\sep climate change\sep 
clinical trials
\end{keywords}

\setstretch{2}
\section{Introduction}\label{introduction}

Linear projections are useful in many aspects of statistical analysis of
multivariate data, and especially useful for visualising the data. A
linear projection provides a dimension reduction while maintaining
interpretability. For example, a biplot (Gabriel 1971; Gower and Hand
1996) shows the structure creating the maximum variance in the data, and
also visualizing the projection matrix to learn which variables
contribute to it. We might find clusters of outliers that were hiding in
high dimensions.

More generally, projection pursuit (Friedman and Tukey 1974; Jones and
Sibson 1987; Huber 1985) defines a quantitative criterion for the
interestingness of a projection (a projection pursuit index), and
searches the space of possible projections for the most interesting one
to display. We can also define sequences of interpolated linear
projections to better understand a multivariate distribution. Animating
a randomly selected interpolated sequence of linear projections is
called a grand tour (Asimov 1985; Buja and Asimov 1986; Buja et al.
2005; D. Cook et al. 2006; Lee et al. 2022). The combination of these
two approaches would then use a projection pursuit index to select
interesting projections, but display them via an interpolated path to
provide context. This is called a guided tour (Dianne Cook et al. 1995).

@@

To define a measure of an interesting projection is to maximize the
average Mahahlanobis distance (Mahalanobis 1936) in the projection for a
subset of points, \(W\). The set of points could be chosen in different
ways, but the default is those that are outside the specified ellipsoid
in \pD{}. Alternatives could be to select a set of observations with the
largest Mahalanobis distance, manually select observations or possibly a
group of points identified via clustering of the extremes.

@@

This is implemented in the \texttt{tourr} (Hadley Wickham et al. 2011;
H. Wickham et al. 2024) package, where the projected ellipsoid can be
drawn for each projection. The guided tour will take arguments
specifying the data, and the null variance-covariance matrix.

@@

This example is motivated by a problem posed during consulting with a
pharmaceutical company, but the data shown here is simulated, simply to
illustrate the application. Liver function tests commonly provide
measurements on albumin, protein, bilirubin, gamma-glutamyl- transferase
(GGT), aspartate aminotransferase (AST), alkaline phosphatase (ALP) and
alanine aminotransferase (ALT). There are normal ranges on these
measurements reported by Lala, Zubair, and Minter (2023) and listed in
Table~\ref{tbl-liver-norms}.

@@

Figure~\ref{fig-liver} illustrates two examples. The first is similar to
the consulting project. A sample of liver test scores for new patients
was provided in order to examine their values relative to the normal
range. Here only four tests are used, GGT, AST, ALP, ALT. Plot (a) shows
a projection of this sample relative to the normal range. Three of the
patients are outside the confidence ellipse but all of the patients are
located away from the mean. What has been typical in the past is to
compute normal values based on tests of healthy young males. The samples
provided for the project were all recorded on women. We see that this
sample has slightly lower ALT and ALP, which is consistent with what is
reported in Lala, Zubair, and Minter (2023).

@@

This example is motivated by the weather data example from Filzmoser,
Hron, and Templ (2018). We illustrate using the anomaly index to compare
potential outliers with a reference normal distribution that is derived
using robust methods on the original data. The data contains 16
numerical measurements that are averages across the three summer months
June, July and August, reported for 68 years (1955 - 2022) (see
Filzmoser, Hron, and Templ (2018) for more details).

To estimate the underlying normal distribution the data is first
centered and scaled using the median and the median absolute deviation
(MAD), before applying the minimum covariance determinant (MCD)
estimator (Peter J. Rousseeuw 1985) using the implementation in Maechler
et al. (2024). The MCD estimates for the mean vector and the
variance-covariance matrix are then used to define the reference normal
distribution.

Here we will consider points to be outlying if they are more than
\(5 \sigma\) away from that mean value. With \(p=16\) this corresponds
to a Mahalanobis distance of \(60\) or larger. This will identify \(20\)
out of the \(68\) observations as outlying. Since these are outlying in
different combinations of variables, as found in Filzmoser, Hron, and
Templ (2018), we will further separate the outlying points into clusters
based on similarity in direction. The similarity is computed by first
normalizing observations to have length \(1\) and then apply k-means
clustering with Euclidean distance. We use the Dunn index Halkidi,
Batistakis, and Vazirgiannis (2002), computed using Hennig (2024), to
select the preferred number of clusters, \(k=5\).

@@

To compare our results to what is obtained by outlier detection methods,
we reproduce one of the results from Filzmoser, Hron, and Templ (2018)
in Figure~\ref{fig-shapley}. On top we can see the cluster assignments
for all points that we have identified as outliers using the
\(p\)-dimensional Mahalanobis distance.

For example we see that cluster 4 contains the years 1962 and 1996, and
for these two years the result from Filzmoser, Hron, and Templ (2018)
also indicates unusually low values for the minimum teperature, while
there is disagreement between outlyingness in other variables. Note that
compared to the years 1965 and 1966 they have unusually low minimum
temperature while having average values for the number of days with
maximum temperature above 25°C, and this contrast is also captured in
the projection.

@@

\caption{\label{fig-shapley}Reproduction of the results from Filzmoser,
Hron, and Templ (2018), showing which cells have been identified as
outlying by their Shapley Cell Detector algorithm. Red cells indicate
values higher than expected, blue cells values lower than expected. For
years that have been identified as outliers in our approach we show the
cluster assignment at the top.}

@@

The work is related to outlier detection methods (see, Filzmoser, Hron,
and Templ 2018, for example), particularly those that use robust
statistics, as illustrated in the weather example. Methods discussed in
Peter J. Rousseeuw, Raymaekers, and Hubert (2018) and Donoho and Gasko
(1992) describe detecting outlyingness using projections of the data.
When these projections are collected and used as a large set, they
define a high-dimensional confidence region of an indeterminate shape.
There is no analytical definition. However, we could explore the data
relative to the high-dimensional shapes using the same approach
described in Section~\ref{sec-background}. You would generate points on
the surface of the irregular shape, overlay the data on this to
visualise with a tour.

@@

Another potential direction of this work is with model-based clustering
using Gaussian mixtures (Fraley and Raftery 2002), where ellipses form
the basis of the model. Currently, the \texttt{mclust} software (Scrucca
et al. 2023) has the capacity to show \gD{} ellipses for two variables,
or axis parallel \gD{} ellipses for \pD{}. The equations developed here
for drawing the \gD{} projection of a \pD{} ellipse would provide more
versatile visualisation for examining the model-based clustering fits.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-As85}
Asimov, D. 1985. {``{T}he {G}rand {T}our: {A} {T}ool for {V}iewing
{M}ultidimensional {D}ata.''} \emph{SIAM Journal of Scientific and
Statistical Computing} 6 (1): 128--43.

\bibitem[\citeproctext]{ref-BA86b}
Buja, A., and D. Asimov. 1986. {``{G}rand {T}our {M}ethods: {A}n
{O}utline.''} \emph{Computing Science and Statistics} 17: 63--67.

\bibitem[\citeproctext]{ref-BCAH05}
Buja, A., D. Cook, D. Asimov, and C. Hurley. 2005. {``{C}omputational
{M}ethods for {H}igh-{D}imensional {R}otations in {D}ata
{V}isualization.''} In \emph{Handbook of Statistics: Data Mining and
Visualization}, edited by C. R. Rao, E. J. Wegman, and J. L. Solka,
391--414. Amsterdam, The Netherlands: Elsevier/North-Holland.

\bibitem[\citeproctext]{ref-cook1995}
Cook, Dianne, Andreas Buja, Javier Cabrera, and Catherine Hurley. 1995.
{``Grand Tour and Projection Pursuit.''} \emph{Journal of Computational
and Graphical Statistics} 4 (3): 155--72.
\url{https://doi.org/10.1080/10618600.1995.10474674}.

\bibitem[\citeproctext]{ref-CLBW06}
Cook, D., E.-K. Lee, A. Buja, and H. Wickham. 2006. {``{G}rand {T}ours,
{P}rojection {P}ursuit {G}uided {T}ours and {M}anual {C}ontrols.''} In
\emph{Handbook of {D}ata {V}isualization}, edited by C.-H. Chen, W.
Härdle, and A. Unwin. Berlin: Springer.

\bibitem[\citeproctext]{ref-donoho1982}
Donoho, David L., and Miriam Gasko. 1992. {``{Breakdown Properties of
Location Estimates Based on Halfspace Depth and Projected
Outlyingness}.''} \emph{The Annals of Statistics} 20 (4): 1803--27.
\url{https://doi.org/10.1214/aos/1176348890}.

\bibitem[\citeproctext]{ref-filzmoser2018}
Filzmoser, P., K. Hron, and M Templ. 2018. \emph{Applied Compositional
Data Analysis}. Cham, Switzerland: Springer.
\url{https://doi.org/10.1007/978-3-319-96422-5}.

\bibitem[\citeproctext]{ref-FR02}
Fraley, C., and A. E. Raftery. 2002. {``Model-Based {C}lustering,
{D}iscriminant {A}nalysis, {D}ensity {E}stimation.''} \emph{Journal of
the American Statistical Association} 97: 611--31.

\bibitem[\citeproctext]{ref-FT74}
Friedman, J. H., and J. W. Tukey. 1974. {``{A} {P}rojection {P}ursuit
{A}lgorithm for {E}xploratory {D}ata {A}nalysis.''} \emph{IEEE
Transactions on Computing C} 23: 881--89.

\bibitem[\citeproctext]{ref-Ga71}
Gabriel, K. R. 1971. {``The {B}iplot {G}raphical {D}isplay of {M}atrices
with {A}pplications to {P}rincipal {C}omponent {A}nalysis.''}
\emph{Biometrika} 58: 453--67.

\bibitem[\citeproctext]{ref-GH96}
Gower, J. C., and D. J. Hand. 1996. \emph{Biplots}. London: Chapman;
Hall.

\bibitem[\citeproctext]{ref-dunn}
Halkidi, Maria, Yannis Batistakis, and Michalis Vazirgiannis. 2002.
{``Clustering Validity Checking Methods: Part II.''} \emph{ACM SIGMOD
Record} 31 (September). \url{https://doi.org/10.1145/601858.601862}.

\bibitem[\citeproctext]{ref-fpc}
Hennig, Christian. 2024. \emph{Fpc: Flexible Procedures for Clustering}.
\url{https://CRAN.R-project.org/package=fpc}.

\bibitem[\citeproctext]{ref-Hu85}
Huber, P. J. 1985. {``{P}rojection {P}ursuit (with Discussion).''}
\emph{Annals of Statistics} 13: 435--525.

\bibitem[\citeproctext]{ref-JS87}
Jones, M. C., and R. Sibson. 1987. {``{W}hat Is {P}rojection {P}ursuit?
(With Discussion).''} \emph{Journal of the Royal Statistical Society,
Series A} 150: 1--36.

\bibitem[\citeproctext]{ref-lib-med-liver-norms}
Lala, V., M. Zubair, and D. A. Minter. 2023. {``Liver Function Tests.''}
\url{https://www.ncbi.nlm.nih.gov/books/NBK482489/}.

\bibitem[\citeproctext]{ref-tours2022}
Lee, Stuart, Dianne Cook, Natalia da Silva, Ursula Laa, Nicholas
Spyrison, Earo Wang, and H. Sherry Zhang. 2022. {``The State-of-the-Art
on Tours for Dynamic Visualization of High-Dimensional Data.''}
\emph{WIREs Computational Statistics} 14 (4): e1573.
\url{https://doi.org/10.1002/wics.1573}.

\bibitem[\citeproctext]{ref-robustbase}
Maechler, Martin, Peter Rousseeuw, Christophe Croux, Valentin Todorov,
Andreas Ruckstuhl, Matias Salibian-Barrera, Tobias Verbeke, Manuel
Koller, Eduardo L. T. Conceicao, and Maria Anna di Palma. 2024.
\emph{Robustbase: Basic Robust Statistics}.
\url{http://robustbase.r-forge.r-project.org/}.

\bibitem[\citeproctext]{ref-mahalanobis}
Mahalanobis, P. C. 1936. {``On the Generalised Distance in
Statistics.''} \emph{Sankhya A} 80 (Suppl 1): 1--7.
\url{https://doi.org/10.1007/s13171-019-00164-5}.

\bibitem[\citeproctext]{ref-rousseeuw1985multivariate}
Rousseeuw, Peter J. 1985. {``Multivariate Estimation with High Breakdown
Point.''} \emph{Mathematical Statistics and Applications} 8 (283-297):
37.

\bibitem[\citeproctext]{ref-directional-outlyingness}
Rousseeuw, Peter J., Jakob Raymaekers, and Mia Hubert. 2018. {``A
Measure of Directional Outlyingness with Applications to Image Data and
Video.''} \emph{Journal of Computational and Graphical Statistics} 27
(2): 345--59. \url{https://doi.org/10.1080/10618600.2017.1366912}.

\bibitem[\citeproctext]{ref-mclust}
Scrucca, Luca, Chris Fraley, T. Brendan Murphy, and Adrian E. Raftery.
2023. \emph{Model-Based Clustering, Classification, and Density
Estimation Using {mclust} in {R}}. Chapman; Hall/CRC.
\url{https://doi.org/10.1201/9781003277965}.
